---
title: "Function documentation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

### c()

Combine values into a vector

```{r eval=FALSE}
c(...)
```

`...` - any number of values

```{r}
x <- c(1, 5, 7, 4, 77, 3, "a")
x
```

### list()

Creates a list object out of multiple vectors

```{r eval=FALSE}
list(...)
```

`...` - any number of vectors

```{r}
l <- list(c(1, 2, 3), c("a", "b", "c"), c(1, "a", 3, "b"))
l
```

### data.frame()

Creates a dataframe object when supplied with column names and values. Each column must have the same number of values

```{r eval=FALSE}
data.frame(…)
```

`...` - one or more name and value pairs, in the form `name=values`

```{r}
df <- data.frame(fruit = c("apple", "orange", "banana"), color = c("red", "orange", "yellow"))
df
```

### head()

displays the first `n` rows of a data frame, or the first `n` values of a list or vector

```{r eval=FALSE}
head(x, n)
```

`x` - data frame, vector, or list\
`n` - number of values to return

```{r}
#df <- data.frame(fruit = c("apple", "orange", "banana"), color = c("red", "orange", "yellow"))
#head(df, 2)

head(penguins, 10)
```

### tail()

displays the last `n` rows of a data frame, or the last `n` values of a list or vector

```{r eval=FALSE}
tail(x, n)
```

`x` - data frame, vector, or list\
`n` - number of values to return

```{r}
#df <- data.frame(fruit = c("apple", "orange", "banana"), color = c("red", "orange", "yellow"))
#tail(df, 2)
tail(penguins, 10)
```

### str()

Displays the internal structure of an object

```{r eval=FALSE}
str(object)
```

`object` - any R object you'd like information about

```{r}
l <- list(c(1, 2, 3), c("a", "b", "c"), c(1, "a", 3, "b"))
str(l)
```

### summary()

Reports summary statistics for a given object. Useful for reporting information on data and models.

```{r eval=FALSE}
summary(object)
```

`object` - any R object you'd like to summarize

```{r}
summary(penguins)
```

### table()

Uses the cross-classifying factors to build a contingency table of the counts at each combination of factor levels. Must include useNA option

```{r eval=FALSE}
table(..., useNA = c("no", "ifany", "always"))
```

`...` - data to summarize with a contingency table\
`useNA` - how to handle NA values

```{r}
table(penguins$species, penguins$island, useNA="no")
```

### install.packages()

Install an R package

```{r eval=FALSE}
install.packags(package)
```

`package` - name of package to be installed, in quotes

```{r eval=FALSE}
install.packages("dplyr")
```

### library()

Load an R package

```{r eval=FALSE}
library(package)
```

`package` - name of package to load

```{r}
library(dplyr)
```

### read_csv()

Read a csv file into a tibble

```{r eval=FALSE}
read_csv(filename, sep = ",")
```

`filename`- name/path to the desired file\
`sep` - value delimiter used in file (usually a comma)

```{r eval=FALSE}
read_csv("data.csv")
```

### read_dta()

Read a Stata file into a tibble

```{r eval=FALSE}
read_csv(filename)
```

`filename`- name/path to the desired file

```{r eval=FALSE}
read_csv("data.dta")
```

### read_sav()

Read a SPSS file into a tibble

```{r eval=FALSE}
read_csv(filename)
```

`filename`- name/path to the desired file

```{r eval=FALSE}
read_csv("data.dta")
```

### mean()

Calculates the arithmetic mean of a vector or column. Must include na.rm option.

```{r eval=FALSE}
mean(x, na.rm = FALSE)
```

`x`- data to calculate the mean of\
`na.rm` - TRUE or FALSE, whether to include NA values or not

```{r}
mean(penguins$bill_length_mm, na.rm = TRUE)
```

### filter()

Selects only rows from the data frame that satisfy the given criteria

```{r eval=FALSE}
filter(.data, ...)
```

`.data`- data frame to filter (unless piped)\
`...` - at least one conditional statement to filter with

```{r}
filter(penguins, species=="Gentoo")

penguins %>% 
  filter(species=="Gentoo")
```

### is.na()

Find NA values in a vector

```{r eval=FALSE}
is.na(x)
```

`x` - vector we'd like to find NAs in

```{r}
is.na(penguins$sex)
```

### write_csv()

Write a data frame to a .csv file

```{r eval=FALSE}
write_csv(df, path)
```

`df` - data frame to save\
`path` - path to new file

```{r eval=FALSE}
write_csv(penguins, "data/penguins.csv")
```

### save() - write objects to a .RData file ???

### count()

Count unique occurrences in a data frame

```{r eval=FALSE}
count(df, vars)
```

`df` - data frame\
`vars` - variables to count unique values of

```{r}
count(penguins, vars=island)
```

mutate() Add a new column to a dataframe or change an existing column

```{r eval=FALSE}
mutate(.data, ...)
```

`df` - data frame to change (unless piped)\
`...` - key value pairs of `name=vector`

```{r}
mutate(penguins, bill_ratio=bill_length_mm/bill_depth_mm)

penguins %>% 
  mutate(bill_ratio=bill_length_mm/bill_depth_mm)
```

### replace_na()

Replace NA values in a data frame with a given value

```{r eval=FALSE}
if_na(x, value)
```

`x` - vector to replace NA values in\
`value` - list of column value pairs to replace NA values with

```{r}
penguins %>% 
  replace_na(list(bill_depth_mm = 0, bill_length_mm = 0))
```

### if_else()

Conditional statement, if true, return one value, otherwise a second value.

```{r eval=FALSE}
if_else(condition, true, false, missing = NULL)
```

`condition` - true/false logical statement\
`true` - value to return if true\
`false` - value to return if false\
`missing`- value to return if data is missing

```{r}
if_else(penguins$bill_depth_mm > 18, "long", "short", missing = "missing")
```

### case_when()

Conditional statements, takes any number of conditions and evaluates in order

```{r eval=FALSE}
case_when(...)
```

`...` - one or more pairs of conditions and values

```{r}
case_when(penguins$bill_depth_mm > 18 ~ "long",
          penguins$bill_depth_mm > 20 ~ "very long",
          penguins$bill_depth_mm > 15 ~ "short",
          penguins$bill_depth_mm > 10 ~ "very short",
          TRUE ~ "other"
          )
```

### as.numeric()

Convert object to numeric values

```{r eval=FALSE}
as.numeric(x)
```

`x` - vector to convert

```{r}
x <- c('1', '2', '3')
as.numeric(x)
```

as.character() Convert object to string values

```{r eval=FALSE}
as.character(x)
```

`x` - vector to convert

```{r}
x <- c(1, 2, 3)
as.character(x)
```

### rename()

Rename one or more columns in a data frame

```{r eval=FALSE}
rename(.data, ...)
```

`.data`- data frame to rename columns of (unless piped)\
`...` - one or more new=old column name pairs

```{r}
rename(penguins, mass = body_mass_g, flipper = flipper_length_mm)

penguins %>% 
  rename(mass = body_mass_g, flipper = flipper_length_mm)
```

### select()

Select one or more columns of a data frame

```{r eval=FALSE}
select(.data, ...)
```

`.data`- data frame to select columns of (unless piped)\
`...` - one or more column names

```{r}
select(penguins, island, species)

penguins %>% 
  select(island, species)
```

### hist()

Plot a histogram, given a vector of data

```{r eval=FALSE}
hist(x)
```

`x` - data to plot

```{r}
hist(penguins$bill_length_mm)
```

### summarize()

Generates a dataframe displaying the results of chosen functions on the data. Often used with group_by().

```{r eval=FALSE}
summarize(.data, ...)
```

`.data`- data frame to summarize (unless piped)\
`...` - one or more summary methods of the form `name=method(x)`

```{r}
summarize(penguins, mean_bill=mean(bill_length_mm, na.rm=TRUE), sd_bill=sd(bill_length_mm, na.rm=TRUE))

penguins %>% 
  group_by(species) %>% 
  summarize(mean_bill=mean(bill_length_mm, na.rm=TRUE),sd_bill=sd(bill_length_mm, na.rm=TRUE))
```

### group_by()

Often used with summarize(), groups the data in the data frame based on one or more columns.

```{r eval=FALSE}
group_by(.data, ...)
```

`.data`- data frame to summarize (unless piped)\
`...` - one or more columns to group by, in order

```{r}
penguins %>% 
  group_by(island, species) %>% 
  summarize(mean_bill=mean(bill_length_mm, na.rm=TRUE),sd_bill=sd(bill_length_mm, na.rm=TRUE))
```

### left_join()

Joins dataframe x with dataframe y, based on the rows/key in x. Rows present in y and not present in x will be omitted. Rows present in x but not y will be kept.

```{r eval=FALSE}
left_join(x,y)
```

`x` - data to join `y` - data to join to x

```{r}

```
